### 2016-01-25 inputs/discussion.tex
aenderungen-19:55

\subsection{The role of the crop species: rice vs. alfalfa}
\label{sec:role_of_species}
Imputation results were significantly different in rice and alfalfa: all imputation algorithms performed consistently better in rice data sets, where Beagle, RFI and KNNI achieved performances comparable to what is reported in literature for SNP-chip data (e.g.,~\cite{vanraden_genomic_2013} for bovine data;
%problems with this citation: url contains hashes
\cite{the_1000_genomes_project_consortium_integrated_2012}
for human data). On the other hand, imputation accuracy in alfalfa data sets was much lower. This can be ascribed to three main factors: 
i) the imputation problem was simpler in rice than in alfalfa, as the rice dataset comprised only pure lines, i.e. only two genotype classes instead of alfalfa's three; 
ii) the genome structure was simpler in rice (diploid) than alfalfa (autotetraploid, rendered diploid ``in silico'' during SNP calling);
iii) rice data were analyzed by chromosome, allowing for higher average values of linkage disequilibrium (LD) among SNP markers, thus simplifying the imputation process.\\
Finally, the further difference in imputation accuracy between Alfalfa-Med and Alfalfa-PV suggests that other factors, directly connected to GBS protocols, may influence the general quality of the resulting data.

@@

\subsection{Size of the problem}
\label{sec:size_of_problem}
The size of the problem, besides affecting the accuracy achievable by some imputation methods, is mainly relevant with reference to the required computation resources. RFI was the most demanding algorithm, with computation times growing exponentially both with the total number of marker genotypes (m SNP $\times$ n samples) and with the number of missing genotypes to be imputed. When analyzing the complete rice dataset (all chromosomes together), RFI took a maximum of about 40 days to complete imputation, and became computationally intractable (on the available platform) for the largest missing rate scenarios. This was partially mitigated by parallelization (we used 10 CPUs in our experiments). All other imputation algorithms took much shorter times to complete imputation, even in the most challenging scenarios. Putting together imputation accuracy and computation time, we found that the best performing imputation algorithms were Beagle in rice and KNNI in alfalfa.\\
The analysis of the rice complete data set (437 samples $\times$ 166\,418 SNP markers) offered interesting insights on the applicability of imputation methods. While problems related to the computational requirements can usually be tackled by the constant improvement in bioinformatics facilities, current trends in genotyping suggest that imputation methods and computational strategies well suited for large data sets are to be preferred. The non-linearity of computational efficiency generated, for most examined imputation algorithms, a total computation time for the complete rice data set which was greater than the sum of the computation times required by the individual chromosomes. Again, RFI provided the most blatant illustration of this behavior, with computation times for the complete rice data set going up to 937 hours (about 40 days), while the sum of the computation times for all the single chromosomes in the same experimental conditions added to 88.3 hours.

>>>>

\subsection{Imputation efficiency differences in rice vs. alfalfa}
\label{sec:role_of_species}
Imputation results were significantly different in rice and alfalfa: all imputation algorithms performed consistently better in rice data sets, where Beagle, RFI and KNNI achieved performances comparable to what is reported in literature for SNP-chip data (e.g.,~\cite{vanraden_genomic_2013} for bovine data;
%problems with this citation: url contains hashes
\cite{the_1000_genomes_project_consortium_integrated_2012}
for human data). On the other hand, imputation accuracy in alfalfa data sets was much lower. This can be ascribed to three main factors: 
i) the imputation problem was simpler in rice than in alfalfa, as the rice dataset comprised only pure lines, i.e. only two genotype classes instead of alfalfa's three; 
ii) the genome structure was simpler in rice (diploid) than alfalfa (autotetraploid, rendered diploid ``in silico'' during SNP calling);
iii) rice data were analyzed by chromosome, allowing for higher average values of linkage disequilibrium (LD) among SNP markers, thus simplifying the imputation process.\\
Finally, the further difference in imputation accuracy between Alfalfa-Med and Alfalfa-PV suggests that other factors, directly connected to GBS protocols, may influence the general quality of the resulting data.

@@

\subsection{Size of the imputation problem}
\label{sec:size_of_problem}
Albeit this was not the main objective of the paper, and no formal effort to optimise the implementation of the various imputation methods was done, still the recorded computation times give us some interesting information.
The size of the problem, besides affecting the accuracy achievable by some imputation methods, is mainly relevant with reference to the required computation resources. RFI was the most demanding algorithm, with computation times growing exponentially both with the total number of marker genotypes (m SNP $\times$ n samples) and with the number of missing genotypes to be imputed. When analyzing the complete rice dataset (all chromosomes together), RFI took a maximum of about 40 days to complete imputation, and became computationally intractable (on the available platform) for the largest missing rate scenarios. This was partially mitigated by parallelization (we used 10 CPUs in our experiments). All other imputation algorithms took much shorter times to complete imputation, even in the most challenging scenarios. Putting together imputation accuracy and computation time, we found that the best performing imputation algorithms were Beagle in rice and KNNI in alfalfa.\\
The analysis of the rice complete data set (437 samples $\times$ 166\,418 SNP markers) offered interesting insights on the applicability of imputation methods. While problems related to the computational requirements can usually be tackled by the constant improvement in bioinformatics facilities, current trends in genotyping suggest that imputation methods and computational strategies well suited for large data sets are to be preferred. The non-linearity of computational efficiency generated, for most examined imputation algorithms, a total computation time for the complete rice data set which was greater than the sum of the computation times required by the individual chromosomes. Again, RFI provided the most blatant illustration of this behavior, with computation times for the complete rice data set going up to 937 hours (about 40 days), while the sum of the computation times for all the single chromosomes in the same experimental conditions added to 88.3 hours.

### 2016-01-25 inputs/introduction.tex
aenderungen-19:55

as high as 0.84-0.94 (measured as $R^{2}$ between true and imputed genotypes). Imputation of missing GBS data without alignment to the reference genome has been reported in alfalfa (``Medicago sativa'', \cite{Rocher_validation_2015}) and red raspberry (``Rubus idaeus'', \cite{ward2013}).\\

>>>>

as high as 0.84-0.94 (measured as $R^{2}$ between true and imputed genotypes). Imputation of missing GBS data without alignment to the reference genome has been reported in alfalfa (``Medicago sativa'', \cite{Rocher_validation_2015}) and red raspberry (``Rubus idaeus'', \cite{ward_2013_saturated}).\\

### 2016-01-25 inputs/materials_and_methods.tex
aenderungen-19:55

\subsection{Imputation methods}
\label{sec:imputation_methods}
We considered five imputation methods: Mean Value Imputation (MNI), K-Nearest Neighbour Imputation (KNNI), Random Forest Imputation (RFI), Singular Value Decomposition Imputation (SVDI), and Localized Haplotype Clustering Imputation (``Beagle''). For all algorithms, we imputed a $m\times n$ matrix of m individuals and n markers whose data points, defined in the set \{0,1,2,NA\}, represented the three possible genotypes (AA, AB, and BB) and the missing value, respectively. 
\paragraph{MNI}
\label{par:MNI}
: in Mean Value Imputation each missing data point was simply replaced by the mean of the non-missing values for that marker, which was then discretized to the closest value in \{0,1,2\}. 
\paragraph{KNNI}
\label{par:KNNI}
: in K-Nearest Neighbors Imputation, missing data points were imputed based on the weighted average of the K closest markers \cite{troyanskaya_missing_2001}. This K-neighbourhood was defined by the Simple Matching Coefficient distance function \cite{schwender_statistical_2007}, specifically designed for categorical data. K=4 was used in KNNI, and neighbour values were weighted by the reciprocal of their distance from the data point to be imputed.
\paragraph{SVDI}
\label{par:SVDI}
: Singular Value Decomposition (SVD) imputation was based on the following factorization of the genotype matrix M (m markers, n individuals):

@@

\begin{equation}
\label{eq:SVDI_regression}
Y = \hat{U}\beta +\varepsilon 
\end{equation}

@@

\paragraph{Beagle}
\label{par:Beagle}
: ``localized haplotype clustering imputation'' is a method implemented in the software ``Beagle'' \cite{browning_rapid_2007}. Originally developed for human genetics, Beagle has since found wide application also in animal and plant genetics. Beagle has become so popular that the name of the software is commonly used  to refer metonymically to the method it implements, making the two hardly distinguishable. Beagle infers haplotypes and imputes sporadic missing alleles both with known and unknown phase, using a localized haplotype cluster model; this is a class of directed acyclic graphs which empirically models haplotype frequencies on a local scale and therefore adapts to local structure in the data. Beagle makes use of a hidden Markov model to find the most likely haplotype pair for each individual, given the genotype data for that individual and the graphical haplotype frequency model. The method works iteratively using an expectation-maximization (EM) approach. The imputed missing data, probabilities of missing genotypes and inferred haplotypes are calculated from the model that is fitted in the last iteration. For the imputation of missing genotypes in the completely homozygous rice accessions, a likelihood file was defined to specifically exclude the imputation of non-possible heterozygous genotypes.

@@

\subsection{Imputation accuracy and efficiency assessment}
\label{sec:imputation_accuracy_and_efficiency_assessment}
We used the artificial missing genotypes to measure the performance of the five imputation methods. The rice dataset contained only two genotype classes: AA (homozygous for the major allele) and BB (homozygous for the minor allele). Both alfalfa datasets contained also the third genotype classes AB (heterozygous, pooling the three possible heterozygotes AAAB, AABB, ABBB).\\
For each experiment we measured the overall imputation accuracy and the imputation accuracy within each genotype class. The imputation accuracy was computed as the number of missing data correctly imputed divided by the total number of missing data (proportion of correct imputations):

\begin{equation}
\label{eq:accuracy}
accuracy = \frac{1}{n} \sum_{i=1}^{n}{I(y_i = \hat{y_i})}
\end{equation}

where $I()$ is an indicator function that returns 1 if the imputed ($\hat{y}$) and true ($y$) genotypes are equal, 0 otherwise. We obtained four accuracy measures for alfalfa (three genotype classes plus total accuracy) and three for rice (two genotype classes plus total accuracy).\\
Imputation methods were also compared in terms of computation efficiency, measured as the total computation time needed to complete the analyses. The measured time equals the total dedicated CPU time in case of single thread execution, and corresponds to a fraction of it in case of algorithm supporting multi thread execution. In our experiments only RFI implemented parallel execution, and it was configured to use 10 CPUs, thus all the reported times are to be considered as time elapsed while employing ten times more computational resources than the other algorithms.\\
To ensure consistency in the experimental conditions, all analysis were run on the same computing platform: a multi-node cluster with 64 CPUs AMD Opteron(tm) 2400 MHz and 256GB RAM for each node.

>>>>

\subsection{Imputation methods}
\label{sec:imputation_methods}
We considered six imputation methods: Mean Value Imputation (MNI), K-Nearest Neighbour Imputation (KNNI), Random Forest Imputation (RFI), Singular Value Decomposition Imputation (SVDI), Localized Haplotype Clustering Imputation (``Beagle'') and the ``FILLIN'' algorithm from the ``Tassel'' programming suite \cite{swarts_novel_2014}. For all algorithms, we imputed a $m\times n$ matrix of m individuals and n markers whose data points, defined in the set \{0,1,2,NA\}, represented the three possible genotypes (AA, AB, and BB) and the missing value, respectively. 
\paragraph{MNI}
\label{par:MNI}
: in Mean Value Imputation each missing data point was simply replaced by the mean of the non-missing values for that marker, which was then discretized to the closest value in \{0,1,2\}. 
\paragraph{KNNI}
\label{par:KNNI}
: in K-Nearest Neighbors Imputation, missing data points were imputed based on the weighted average of the K closest markers \cite{troyanskaya_missing_2001}. This K-neighbourhood was defined by the Simple Matching Coefficient distance function \cite{schwender_statistical_2007}, specifically designed for categorical data. $K=4$ was used in KNNI, and neighbour values were weighted by the reciprocal of their distance from the data point to be imputed.
\paragraph{SVDI}
\label{par:SVDI}
: Singular Value Decomposition (SVD) imputation was based on the following factorization of the genotype matrix M (m markers, n individuals):

@@

\begin{equation}
\label{eq:SVDI_regression}
Y = U\beta +\varepsilon 
\end{equation}

@@

\paragraph{Beagle}
\label{par:Beagle}
: ``localized haplotype clustering imputation'' is a method implemented in the software ``Beagle'' \cite{browning_rapid_2007}. Originally developed for human genetics, Beagle has since found wide application also in animal and plant genetics. Beagle has become so popular that the name of the software is commonly used  to refer metonymically to the method it implements, making the two hardly distinguishable. Beagle infers haplotypes and imputes sporadic missing alleles both with known and unknown phase, using a localized haplotype cluster model; this is a class of directed acyclic graphs which empirically models haplotype frequencies on a local scale and therefore adapts to local structure in the data. Beagle makes use of a hidden Markov model to find the most likely haplotype pair for each individual, given the genotype data for that individual and the graphical haplotype frequency model. The method works iteratively using an expectation-maximization (EM) approach. The imputed missing data, probabilities of missing genotypes and inferred haplotypes are calculated from the model that is fitted in the last iteration. For the imputation of missing genotypes in the completely homozygous rice accessions, a likelihood file (with the prior likelihood of each of the three possible SNP genotypes) was defined to specifically exclude the imputation of non-possible heterozygous genotypes (details in \cite{beagle3_manual}).

\paragraph{FILLIN}
The ``Fast Inbred Line Library ImputatioN'' (``FILLIN'') is an imputation method optimised for inbred populations implemented in the ``Tassel'' programming suite \cite{swarts_novel_2014}. ``FILLIN'' is based on haplotype reconstruction around recombination break points; haplotypes are then clustered per genotype similarity using the Hamming distance function. This information is eventually used to impute the target locus in an iterative approach that attemps, through a Markov Chain MonteCarlo (MCMC) process, to maximise the likelihood of the observed SNP calls given the unobserved imputed genotypes.

@@

\subsection{Imputation accuracy and efficiency assessment}
\label{sec:imputation_accuracy_and_efficiency_assessment}
We used the artificial missing genotypes to measure the performance of the five imputation methods. The rice dataset contained only two genotype classes: AA (homozygous for the major allele) and BB (homozygous for the minor allele). Both alfalfa datasets contained also the third genotype classes AB (heterozygous, pooling the three possible heterozygotes AAAB, AABB, ABBB).\\
For each experiment we measured the overall imputation accuracy and the imputation accuracy within each genotype class. The imputation accuracy was computed as the number of missing data correctly imputed divided by the total number of artificially missing data (proportion of correct imputations):

\begin{equation}
\label{eq:accuracy}
accuracy = \frac{1}{n} \sum_{i=1}^{n}{I(g_i = \hat{g_i})}
\end{equation}

where $I()$ is an indicator function that returns 1 if the imputed ($\hat{g}$) and true ($g$) genotypes are equal, 0 otherwise. We obtained four accuracy measures for alfalfa (three genotype classes plus total accuracy) and three for rice (two genotype classes plus total accuracy).\\
For each imputation method, also the total computation time needed to complete the analyses was measured, in order to get an idea of their relative performance. The measured time equals the total dedicated CPU time in case of single thread execution, and corresponds to a fraction of it in case of algorithm supporting multi thread execution. In our experiments only RFI implemented parallel execution, and it was configured to use 10 CPUs, thus all the reported times are to be considered as time elapsed while employing ten times more computational resources than the other algorithms.\\
To ensure consistency in the experimental conditions, all analysis were run on the same computing platform: a multi-node cluster with 64 CPUs AMD Opteron(tm) 2400 MHz and 256GB RAM for each node.
